---
title: "Exploring design matrices"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(dplyr)
library(lumi)
library(limma)
library(pander)
library(stringr)
library(reshape2)
library(ggplot2)

load("R_output/duod_expressed.Rdata")
```

There are 8 Ptids and 2 Visits per Ptid (Enrollment and Visit2)

Here is what the design matrix looks like for the design with no interecept,

~0 + Visit + Ptid

Why are there only 7 Ptids shown? 
I think because the first level of each parameter is absorbed into the intercept and the because I am using an additive model, the coefficients for the Visits are assuming that the effect is the same for all patients. See here: According to http://genomicsclass.github.io/book/pages/expressing_design_formula.html, and replace "Visit" with "Diet" and "Ptid" for "Sex"

Rows 15 and 16 have zeros in all columns except for the first two, which I assume means that they are representing the average of the expression for all Ptids for those visits??


Because there is information on both Visits and I want to compare the average results from the two visits, I will make a contrasts matrix:

contrasts <- makeContrasts(Visit2vsEnrollment = VisitVisit2-VisitEnrollment, levels = no_int_design)
  
  
and then fit the model to the contrast matrix.
```{r no intercept}

Ptid <- factor(pData(duod_expressed)$Ptid) # 2 reps of each Ptid because 2 visits

Visit <-  factor(pData(duod_expressed)$Visit) # 2 Visits per Ptid

no_int_design <- model.matrix(~0+Visit + Ptid)

no_int_fit<-lmFit(duod_expressed, design = no_int_design)

contrasts <- makeContrasts(
  Visit2vsEnrollment = VisitVisit2-VisitEnrollment, levels = no_int_design)

no_int_fit2 <- contrasts.fit(contrasts, fit = no_int_fit)

no_int_fit2<-eBayes(no_int_fit2)


no_int_design


fit_results_sep<- function(fit2){
  results <- decideTests(fit2,method="separate", adjust.method="BH",
                      p.value=0.05, lfc=0.5)
resultsDF<-as.data.frame(results)
resultsDF$ProbeID<-rownames(resultsDF)
rownames(resultsDF)<-NULL

#melt the df for easy summarizing

resultsDFmelt<-melt(resultsDF, id.vars="ProbeID")

summary<-resultsDFmelt %>%
  group_by(variable)%>%
 summarize(down=sum(value=="-1"),up=sum(value=="1"))

pander(summary)
}



fit_results_glob<- function(fit2){
  results <- decideTests(fit2,method="global", adjust.method="BH",
                      p.value=0.05, lfc=0.5)
resultsDF<-as.data.frame(results)
resultsDF$ProbeID<-rownames(resultsDF)
rownames(resultsDF)<-NULL

#melt the df for easy summarizing

resultsDFmelt<-melt(resultsDF, id.vars="ProbeID")

summary<-resultsDFmelt %>%
  group_by(variable)%>%
 summarize(down=sum(value=="-1"),up=sum(value=="1"))

pander(summary)
}
```


There are 8 Ptids and 2 Visits per Ptid (Enrollment and Visit2)

Here is what the design matrix looks like for the design *with* intercept, 
 ~Visit + Ptid


I think the intercept here represents the average of Enrollment and the coefficent "VisitVisit2" represents the increase in the average of Visit2 *over* Enrollment. Because this comparison is already made with this design, I don't need to do a contrast matrix. 

```{r  with intercept}

Ptid <- factor(pData(duod_expressed)$Ptid) # 2 reps of each Ptid because 2 visits

Visit <-  factor(pData(duod_expressed)$Visit) # 2 Visits per Ptid

int_design <- model.matrix(~1 + Visit + Ptid)

int_fit<-lmFit(duod_expressed, design = int_design)


int_fit2<-eBayes(int_fit)


int_design

```

Here are results from fitting the model using decideTests(method = "global") vs decideTests(method = "separate") for the design *without* an intercept.

```{r decide tests no int}
fit_results_sep(no_int_fit2)
fit_results_glob(no_int_fit2)

```



Here are results from fitting the model using decideTests(method = "global") vs decideTests(method = "separate") for the design *with* an intercept.

```{r decide tests int}
fit_results_sep(int_fit2)
fit_results_glob(int_fit2)

```



```{r limma example}

#following section 9.4.1 in the limma guide

# targets<-data.frame("FileName" = c("File1", "File2","File3", "File4","File5","File6"), "SibShip" = c(1,1,2,2,3,3), "Treatment" = c("C","T", "C","T","C","T"))
# SibShip <- factor(targets$SibShip)
# Treat <- factor(targets$Treatment, levels=c("C","T"))
# design <- model.matrix(~SibShip+Treat)
# 
# 
# 
# #this is an example from the edgeR guide
# targets2<-data.frame("FileName" = c("File1", "File2","File3", "File4","File5","File6","File7", "File8", "File9"), "Litter" = c("1","1","1","2","2","2",'3','3',"3"), "Treatment" = c("A","B","C","B","A","C","C","B","A"))
# 
# Litter <- factor(targets2$Litter)
# 
# Treatment <- factor(targets2$Treatment)
# 
# design2 <- model.matrix(~Litter + Treatment)
```