---
title: "PreP Study Microarray Analysis"
author: "Claire Levy"
output: github_document
---

##Experimental set up

We isolated RNA from four different sample types from 8 donors pre- and post- initiation of PreP. The pre-initation was called "Enrollment"  and  the post-initiation visit was called "Visit2"

Sample Types

 * Duodenal biopsy
 * Rectal biopsy
 * PAXgene (whole blood collected into RNA preservative)
 * PBMC (PBMC isolated from whole blood)

NOTE: PTID BG2305 may not have been adherent, left them in for now. May have had problems refilling prescription on time. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(dplyr)
library(lumi)
library(limma)
library(pander)
library(stringr)
library(reshape2)
library(ggplot2)

```

```{r read in raw data, cache = TRUE, results = 'hide'}


lumibatch<-lumiR(fileName = "raw_data/2017.04.18.smhughesFinalReport.txt ",
            detectionTh = 0.05,
            annotationColumn=c('ENTREZ_GENE_ID','ACCESSION', 'SYMBOL', 'PROBE_SEQUENCE', 'PROBE_START', 'CHROMOSOME', 'PROBE_CHR_ORIENTATION', 'PROBE_COORDINATES', 'DEFINITION'))


#read in pData

#Note that my pData file has rownames that match the sampleNames used in the array AND I have a column called sampleNames containing the same data because I'll want to use it later when I make MDS plots and need to merge on sampleNames.  This is annoying but kind of necessary.

pData <-read.table("raw_data/pData.txt", header = TRUE, stringsAsFactors = FALSE)




#put pData into an adf
adf<-AnnotatedDataFrame(data = pData)


complete_lumibatch<-new("LumiBatch",
                     assayData = assayData(lumibatch),
                     phenoData = adf,
                     detection = detection(lumibatch),
                     controlData = controlData(lumibatch),
                     featureData = featureData(lumibatch))
```



Plots of non-normalized data

```{r nonnormalized data}

#MDS

dim_data<- plotMDS(complete_lumibatch)
  d<-data.frame(data.frame(Dim1 = dim_data$x, 
               Dim2 = dim_data$y, 
               sampleNames = names(dim_data$x)))
  
  d<-merge(d, pData, by = "sampleNames")
  
  ggplot(d, aes(x=Dim1, y = Dim2))+
  geom_point(aes(color = Tissue), size = 3)






#A numerical vector of the form c(bottom, left, top, right) which gives the number of lines of margin to be specified on the four sides of the plot. The default is c(5, 4, 4, 2) + 0.1.

par(cex= 0.7, mar = c(10,10,10,10))
boxplot(complete_lumibatch, col = c("#1b9e77","#80cdc1","#984ea3","#e7298a"))

#Ryan Basom looked at this box plot with me and we talked about how it shows that the paxgene samples almost always have low expression and the IQR is smaller. If I normalize all samples together, apparently non-random variation might cause problems. RB's recommendation was to normalize and analyze the tissues separately. 
```


```{r background correction, results = 'hide'}

#the data we got from the core has no background correction (I don't think it did anyway...) so I will do it here.


B_complete_lumibatch<-lumiB(complete_lumibatch, method = "bgAdjust")

```

```{r split up data into tissue types}


duod_batch<-complete_lumibatch[, complete_lumibatch$Tissue =="Duodenal"]

rec_batch<-complete_lumibatch[, complete_lumibatch$Tissue =="Rectal"]

pb_batch<-complete_lumibatch[, complete_lumibatch$Tissue =="PBMC"]


px_batch<-complete_lumibatch[, complete_lumibatch$Tissue =="PAXgene"]

```



```{r vst transform androbust spline normalization, results = 'hide'}


#decided to do the transformation separately for each tissue since the function uses the variance to transform. I get a slightly different number of up and down DE probes than when I transformed all together (together = 76 up, 15 down, separate = 70 up and 11 down)

duod_trans_batch<- lumiT(duod_batch)
duod_norm_batch<-lumiN(duod_trans_batch, method = "rsn")

rec_trans_batch<- lumiT(rec_batch)
rec_norm_batch<-lumiN(rec_trans_batch, method = "rsn")

pb_trans_batch<- lumiT(pb_batch)
pb_norm_batch<-lumiN(pb_trans_batch, method = "rsn")

px_trans_batch<- lumiT(px_batch)
px_norm_batch<-lumiN(px_trans_batch, method = "rsn")

```

These normalized samples all look good.



```{r MDS plot function}

# a function to make nice MDS plots

#plotMDS  invisibly returns a large object that has the actual dimensions data in matrices called x and y (one for each dimension)

#Dim1 is a column containing the values from the x matrix
#Dim2 is a column containing the values from the y matrix
#sampleNames is a column that holds the column names from the x matrix, which happen to be the microarray wells.


#add a plot title to this function!!!


plot_MDS_by_Visit<-function(batch){
  dim_data<- plotMDS(batch)
  d<-data.frame(data.frame(Dim1 = dim_data$x, 
               Dim2 = dim_data$y, 
               sampleNames = names(dim_data$x)))
  
  d<-merge(d, pData, by = "sampleNames")
  
  ggplot(d, aes(x=Dim1, y = Dim2))+
  geom_point(aes(color = Visit), size = 3)
  }

```

```{r norm plots}
#duod
plotDensities(duod_norm_batch,legend = FALSE, main = "Normalized Duodenal Samples")

boxplot(duod_norm_batch,main = "Normalized Duodenal Samples")

plot_MDS_by_Visit(duod_norm_batch)


# rectal
plotDensities(rec_norm_batch,legend = FALSE, main = "Normalized Rectal Samples")

boxplot(rec_norm_batch, main =" Normalized Rectal Samples" )

plot_MDS_by_Visit(rec_norm_batch)

#PBMC
plotDensities(pb_norm_batch,legend = FALSE, main = "Normalized PBMC Samples")

boxplot(pb_norm_batch, main ="Normalized PBMC Samples" )

plot_MDS_by_Visit(pb_norm_batch)


#PAXgene
plotDensities(px_norm_batch,legend = FALSE, main = "Normalized PAXgene Samples")

boxplot(px_norm_batch, main = "Normalized PAXgene Samples" )


plot_MDS_by_Visit(px_norm_batch)
```



## Non-specific filtering
There are  8 donors x 1 tissue type x 2 timepoints = 16 samples in this set.
I will filter out any probes that are not expressed above the 0.05 p-value cut-off in at least 7 samples.  I chose 7 because a probe may not be expressed at all in the 8 Enrollment samples (so 0/16 total), but may show up at Visit2 (8 possibilities). It would still be biologically interesting if a probe was expressed at Visit2 (and not at enrollment) even if it wasn't in all donors, so I'll require it to show up in at least 7 of them.

```{r ns filtering}

#count probes where det pval <0.05 in at least 7 samples
#extract those probes from the batch

find_expressed <-function(batch){
  x <-rowSums(detection(batch)<0.05)>= 7
  batch[x,]
}



duod_expressed<-find_expressed(duod_norm_batch)

rec_expressed<-find_expressed(rec_norm_batch)

pb_expressed<-find_expressed(pb_norm_batch)

px_expressed<-find_expressed(px_norm_batch)
```

Number of probes removed from the data sets after filtering for expression. All started with 47323 probes.

```{r probes pre and post filter}
#here are the probes that were removed by NS filtering for each tissue
probes_removed<- data.frame(
"Duod" = nrow(fData(duod_norm_batch))-nrow(fData(duod_expressed)),
"Rectal" = nrow(fData(rec_norm_batch))-nrow(fData(rec_expressed)),
"PBMC" =nrow(fData(pb_norm_batch))-nrow(fData(pb_expressed)),
"PAXgene" = nrow(fData(px_norm_batch))-nrow(fData(px_expressed)))


pander(probes_removed)
```


```{r design and fit function}


#When I have ~0 for the intercept, each coeff represents the avg of the samples at each level of Condition and Donor. If I  use ~1 instead, the intercept would represent the avg of the samples for the first factor and the next coeff would represent the increase in the average of the next factor OVER the first. See here:https://support.bioconductor.org/p/67395/ and here https://support.bioconductor.org/p/12145/



fit_model<-function(expressedProbes){
  
Ptid <- pData(expressedProbes)$Ptid #get pData just for that tissue

Visit <-  pData(expressedProbes)$Visit

design <- model.matrix(~0 + Visit + Ptid)

fit<-lmFit(expressedProbes, design = design)

contrasts<-makeContrasts(
  Visit2vsEnrollment = VisitVisit2-VisitEnrollment, levels = design)

fit2 <- contrasts.fit(contrasts, fit = fit)

eBayes(fit2)
}

#fit model to each tissue
duod_fit2<-fit_model(duod_expressed)
rec_fit2<-fit_model(rec_expressed)
pb_fit2<-fit_model(pb_expressed)
px_fit2<-fit_model(px_expressed)

```


```{r results summary function}

#results summary function


#method = separate is same as setting the limits for all coefs separately while "global" sets them all at once. I don't think this matters for me since I only have one contrast. I get the same answer if I used global or separate


fit_results<- function(fit2){
  results <- decideTests(fit2,method="global", adjust.method="BH",
                      p.value=0.05, lfc=0.5)
resultsDF<-as.data.frame(results)
resultsDF$ProbeID<-rownames(resultsDF)
rownames(resultsDF)<-NULL

#melt the df for easy summarizing

resultsDFmelt<-melt(resultsDF, id.vars="ProbeID")

summary<-resultsDFmelt %>%
  group_by(variable)%>%
 summarize(down=sum(value=="-1"),up=sum(value=="1"))

pander(summary)
}
  
#summarise each tissue

duod_summary<-fit_results(duod_fit2)
rec_summary<-fit_results(rec_fit2)
pb_summary<-fit_results(pb_fit2)
px_summary<-fit_results(px_fit2)

 
```

##Number of DE probes for each contrast

The duodenal samples were the only ones with any differentially expressed probes. For Visit2 relative to enrollment, there were 70 down-regulated probes and 11 up-regulated probes.


## Changes in the Duodenum pre vs post-PreP: logFC cutoff = 0.5, p-value cut-off = 0.05

```{r toptable}

ttDuod <- topTable(duod_fit2, coef = "Visit2vsEnrollment", adjust.method = "BH", number = Inf, p.value = 0.05, lfc = 0.5 )

pander(ttDuod %>%
         mutate(Direction = ifelse(logFC<0, "DOWN", "UP"))%>%
         select (TargetID, Direction, DEFINITION, adj.P.Val))
        
```

## CAMERA testing

From the CAMERA documentation: 

"camera performs a competitive test in the sense defined by Goeman and Buhlmann (2007). It tests whether the genes in the set are highly ranked in terms of differential expression relative to genes not in the set."

### Hallmark gene sets: top 10 results

```{r camera test setup}

#I'm going to do a CAMERA test using the MSigDB hallmark sets on all of my contrasts
#downloaded Rdata file 22Apr16 from http://bioinf.wehi.edu.au/software/MSigDB/

load("geneSets_for_CAMERA/human_Hallmark_v5.rdata")

#Eset to use

eSet<- exprs(duod_expressed)

#IDs from eSet: entrezIDs from the data I'm analyzing (not the pre-determined gene set). 

identifiers <- as.character(fData(duod_expressed)[,"ENTREZ_GENE_ID"])

```

```{r camera function}

#a function to get the appropriate design and contrast matrices for expressed probes from each tissue, then run camera on them.


camera_function<-function(expressedProbes, geneSet){
  
Ptid <- pData(expressedProbes)$Ptid #get pData just for that tissue

Visit <-  pData(expressedProbes)$Visit

design <- model.matrix(~0+ Visit + Ptid)

contrast <-makeContrasts(
  Visit2vsEnrollment = VisitVisit2-VisitEnrollment, levels = design)

#IDs from eSet that I'll use for camera: entrezIDs *from the data I'm analyzing* (not the pre-determined gene set). 

identifiers <- as.character(fData(expressedProbes)[,"ENTREZ_GENE_ID"])

#make the index of the geneSet using identifiers from my eSet
index <- ids2indices(gene.sets = geneSet, identifiers = identifiers)

camera(y = exprs(expressedProbes), index = index, design = design, contrast = contrast) 

}

```

```{r camera results function}
# a function to make the camera results look better

camera_results<-function(camera_result){
  
camera_result$Gene_Set <- rownames(camera_result)
rownames(camera_result)<-NULL

camera_result%>%
  mutate("Gene_Set" = tolower(Gene_Set))%>%
  head(n = 10)%>%
  mutate(Gene_Set = str_replace(Gene_Set,"hallmark_",""))%>%
  mutate(Gene_Set = str_replace_all(Gene_Set,"_"," "))%>%
  dplyr::select(Gene_Set, Direction, FDR, NGenes)%>%
  pander()
}  
  


```

####Duodenal sample results
```{r hallmark duod}

#I want to use the "Hs.H" hallmark genesets
camera_results(camera_function(duod_expressed, Hs.H))

    
```


####Rectal sample results
```{r hallmark rec}

#I want to use the "Hs.H" hallmark genesets
camera_results(camera_function(rec_expressed, Hs.H))
```

####PBMC sample results
```{r hallmark pb}

#I want to use the "Hs.H" hallmark genesets
camera_results(camera_function(pb_expressed, Hs.H))
```

####PAXgene sample results
```{r hallmark px}

#I want to use the "Hs.H" hallmark genesets
camera_results(camera_function(px_expressed, Hs.H))
```


### GO biological process gene sets: top 10 results

Sean wrote some code (https://github.com/seaaan/Bioinformatics/tree/master/GOTermMappingsForCamera) to extract just the Biological Process gene sets out of all the GO gene sets. 


####Duodenal sample results
```{r }

load("geneSets_for_CAMERA/Hs.c5.BP.rdata")

#I want to use the "Hs.c5.BP"  genesets
camera_results(camera_function(duod_expressed, Hs.c5.BP))

```


####Rectal sample results
```{r }

#I want to use the "Hs.c5.BP"  genesets
camera_results(camera_function(rec_expressed, Hs.c5.BP))
```

####PBMC sample results
```{r }

#I want to use the "Hs.c5.BP"  genesets
camera_results(camera_function(pb_expressed, Hs.c5.BP))
```

####PAXgene sample results
```{r}

#I want to use the "Hs.c5.BP" genesets
camera_results(camera_function(px_expressed, Hs.c5.BP))
```

### MTN-007 gene sets

Here I am comparing the probes in this experiment to significantly differentially expressed probes from MTN-007 9cm biopsies at 7 days.


```{r }

load("geneSets_for_CAMERA/mtn_007_sets.Rda")

#I want to use the "mtn_007_sets"  genesets

mtn_007<-rbind(camera_function(duod_expressed, mtn_007_sets),
      camera_function(rec_expressed, mtn_007_sets),
      camera_function(pb_expressed, mtn_007_sets),
      camera_function(px_expressed, mtn_007_sets))

mtn_007$Gene_Sets<- rownames(mtn_007) 
rownames(mtn_007)<-NULL


mtn_007<-mtn_007 %>%
  mutate(Gene_Sets = rep(c("MTN_007 Down", "MTN_007 Up"), times = 4))%>%
  mutate(Tissue = c("Duodenal","Duodenal", "Rectal","Rectal", "PBMC", "PBMC", "PAXgene", "PAXgene"))%>%
  select(Gene_Sets, Tissue, Direction, FDR, NGenes)


pander(mtn_007)
```


##Which DE probes from the PreP study overlap with those from MTN-007?

```{r mtn007 overlaps}

#read in mtn007 DEG data 7 days 9cm

mtn007_7d_9cm<-read.csv("raw_data/mtn-007-7d-9cm-DEGs.csv")

overlap<-ttDuod[ttDuod$TargetID %in% mtn007_7d_9cm$TargetId,]%>%
  select(TargetID, DEFINITION, logFC)

pander(overlap)

```

## CAMERA test of Hallmark interferon alpha geneset founder sets

```{r inf alpha founder sets }

#to drill down into the hallmark results, I'll run camera on the founder sets of the hallmark interferon alpha set.


library(GSEABase)
library(illuminaHumanv4.db)

#I'm using the illuminaHumanv4.db annotation database because that's the kind of chip we used. http://bioconductor.org/packages/release/data/annotation/html/illuminaHumanv4.db.html  but you can us a different annotation if you want to.

inf_a_founders<-getBroadSets("geneSets_for_CAMERA/HALLMARK_INTERFERON_ALPHA_RESPONSE.xml",geneIdType(EntrezIdentifier(annotation = "illuminaHumanv4.db")))


#map the symbols to entrez IDs because that's the identifier I used in my eSet.
inf_a_founders_entrez<-mapIdentifiers(inf_a_founders, EntrezIdentifier(annotation = "illuminaHumanv4.db"))

#here is my gene set, a list of 82 named elements and in each element are the entrez Ids in the set.

inf_a_founders_set<-geneIds(inf_a_founders_entrez)

```

####Duodenal sample results
```{r test inf alpha}

camera_results(camera_function(duod_expressed, inf_a_founders_set))

```


####Rectal sample results
```{r }
camera_results(camera_function(rec_expressed, inf_a_founders_set))
```

####PBMC sample results
```{r }
camera_results(camera_function(pb_expressed, inf_a_founders_set))
```

####PAXgene sample results
```{r}

camera_results(camera_function(px_expressed, inf_a_founders_set))
```
